---
title: "Why Bioinformatics? Developing Computational Methods to Analyze Biological Data"
author: "Josep Garnica Caparros"
date: "27 March 2026"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)
set.seed(1)
```

**Rendering note**: knitting to HTML requires **Pandoc** (bundled with RStudio, or install via Quarto / system package manager).

# Audience and goals

**Audience**: mostly 3rd-year Biomedical Sciences bachelor students (plus some MSc).

**Goals of this talk**:

- Understand *why* bioinformatics is needed (the “too much data” problem).
- See how computational methods turn data into biological insight.
- Learn the intuition behind a few core ideas:
  - high-dimensional data and the curse of dimensionality
  - dimensionality reduction (PCA) as a tool for interpretation
  - cell type annotation in single-cell RNA-seq
- Understand what it means to *develop* a method/tool (assumptions → algorithm → validation → usability).

# 1. My story: how I ended up doing bioinformatics

(Short personal narrative; adapt live.)

- Started in wet lab work.
- Generated transcriptomic data, which *accumulated faster than we could analyze it*.
- During COVID lockdown: started learning bioinformatics to analyze my own data.
- Realized that good tools matter because they make analysis:
  - reproducible
  - faster
  - more interpretable
  - more accessible for others

If you want a short CV version, see: `docs/CV_JosepGarnicaCaparros_short.txt`.

## The key message

Bioinformatics is not “coding for coding’s sake”. It is **problem-driven**:

- there is a biological question
- data is generated to answer it
- computational methods are needed to extract the answer reliably

# 2. The first bioinformatics “crisis”: too much biological data

A classic example: **protein sequences** in the 1960s–70s.

- As soon as many sequences became available, humans could no longer compare them by eye.
- We needed algorithms to:
  - align sequences
  - detect similarity
  - infer evolutionary relationships
  - identify conserved motifs/domains

## A minimal framing of method development

When you build a computational method, you typically decide:

1. **Input**: what data do we accept?
2. **Representation**: how do we encode it?
3. **Objective**: what do we optimize (similarity, likelihood, classification accuracy)?
4. **Algorithm**: how do we compute it efficiently?
5. **Validation**: how do we know it works?
6. **Packaging**: how do others use it? (docs, defaults, error messages)


# 3. High-dimensional data and the curse of dimensionality


png("figs/height_weight_2d.png", width=600, height=500)

## Simple intuition: height vs weight (2D, with groups)

Suppose you measure just two characteristics of people: height and weight. Here, we simulate two groups (A and B) with different means, and plot them with distinct colors and a legend for clarity, using ggplot2 and ggpubr::theme_classic2().

```{r height-weight-2d, message=FALSE, warning=FALSE, fig.cap="People plotted by height and weight (2D, two groups)", fig.path="figs/", fig.width=6, fig.height=5}
library(ggplot2)
library(ggpubr)
library(patchwork)
set.seed(42)
n <- 100
group <- rep(c("A", "B"), each = n/2)
height <- c(rnorm(n/2, mean = 165, sd = 7), rnorm(n/2, mean = 178, sd = 7))
weight <- c(rnorm(n/2, mean = 62, sd = 8), rnorm(n/2, mean = 80, sd = 10))
df2d <- data.frame(height, weight, group)
cols <- c("#377eb8", "#e41a1c")

p2d <- ggplot(df2d, aes(x = height, y = weight)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = cols, name = "Group") +
  labs(x = "Height (cm)", y = "Weight (kg)",
       title = "People: height vs weight") +
  ggpubr::theme_classic2(base_size = 15)
p2d_color <- ggplot(df2d, aes(x = height, y = weight, color = group)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = cols, name = "Group") +
  labs(x = "Height (cm)", y = "Weight (kg)",
       title = "People: height vs weight (two groups)") +
  ggpubr::theme_classic2(base_size = 15)

```


age <- c(rnorm(n/2, mean = 38, sd = 7), rnorm(n/2, mean = 52, sd = 8))

## Adding more variables: 3D and beyond (with groups)

Now imagine you also measure age and cholesterol. Each person is now a point in 4D space (height, weight, age, cholesterol). Here, we show a 3D plot (height, weight, age) with the same two groups, using color and legend for clarity, using plotly for a modern look.

```{r height-weight-age-3d, message=FALSE, warning=FALSE, fig.cap="People: height, weight, and age (3D, two groups)", fig.path="figs/", fig.width=6, fig.height=5}
library(plotly)
age <- c(rnorm(n/2, mean = 38, sd = 7), rnorm(n/2, mean = 52, sd = 8))
df3d <- data.frame(height, weight, age, group)
cols <- c("#377eb8", "#e41a1c")
p3d <- plot_ly(df3d, x = ~height, y = ~weight, z = ~age, color = ~group, colors = cols, type = "scatter3d", mode = "markers", marker = list(size = 5, opacity = 0.8)) %>%
  layout(scene = list(xaxis = list(title = "Height (cm)"),
                     yaxis = list(title = "Weight (kg)"),
                     zaxis = list(title = "Age (years)")),
         title = "People: height, weight, age (two groups)")
p3d
htmlwidgets::saveWidget(p3d, "figs/height_weight_age_3d.html", selfcontained = TRUE)
```

# 4. PCA in bulk RNA-seq

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(factoextra)
library(DESeq2)
```

## dummy

```{r}
set.seed(22)

# Simulate correlated data
x <- rnorm(100)
y <- x * 0.8 + rnorm(100, sd = 0.3)
df <- data.frame(x, y)

# PCA
pca <- prcomp(df, scale. = TRUE)

# Plot
plot(df$x, df$y, pch = 19, col = "grey",
     main = "PCA finds directions of maximum variance")

# Add PC axes
arrows(0, 0,
       pca$rotation[1,1]*3, pca$rotation[2,1]*3,
       col = "red", lwd = 2)
arrows(0, 0,
       pca$rotation[1,2]*3, pca$rotation[2,2]*3,
       col = "blue", lwd = 2)

legend("topright", legend = c("PC1", "PC2"),
       col = c("red", "blue"), lwd = 2)
```



## Load datasets and process

```{r}

file_path <- "data/GSE173681_Tet_TF_Basal_Stim_Raw_counts.xlsx"

mat_df <- readxl::read_xlsx(file_path)
# process df into matrix

sample_names <- names(mat_df)[7:18] %>% 
  # remove technical replicates
  grep("C_", ., invert = T, value = T)

counts <- mat_df %>% 
  tibble::column_to_rownames("Gene_name") %>% 
  select(all_of(sample_names)) %>% 
  as.matrix()
#Transform the matrix by rounding the counts and transforming the values from "numeric" to "integer"
counts <- apply(counts, c(1,2), round)
mat <- apply(counts, c(1,2), as.integer)
  

metadata <- data.frame(sample_name = sample_names) %>% 
  mutate(condition = ifelse(grepl("Tet", sample_name),
                            "Stimulated",
                            "Basal")
         ) %>% 
  tibble::column_to_rownames("sample_name")
```


Barplots differences between top 10 genes between condition A and cell type B
Barplots differences between top 10 PCA between cell type A and cell type B

## Normaliza data using DESeq2

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = mat,
  colData = metadata,
  design = ~ condition
)

# Remove genes with very low counts
dds <- dds[rowSums(counts(dds)) > 10, ]

# DESeq2 size factor normalization + VST
DESeq.ds_f <- estimateSizeFactors(dds)
vsd <- vst(DESeq.ds_f, blind = TRUE)

norm_counts <- assay(vsd)
```

## Get highly variable genes

```{r}
# Compute variance for each gene across samples
gene_variances <- apply(norm_counts, 1, var)

# Rank genes by variance
gene_variances <- sort(gene_variances, decreasing = TRUE)

# Select top N HVGs (e.g., 2000)
top_n <- 500
hvg_genes <- names(gene_variances)[1:top_n]

# Subset normalized matrix
hvg_matrix <- norm_counts[hvg_genes, ]
```

## barplot hvg

```{r}
tophvg <- names(gene_variances)[100:107] %>% 
  grep("Trav", ., invert = T, value = T)
tophvg_exp <- norm_counts[tophvg,] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("gene") %>% 
  pivot_longer(-gene,
               names_to = "sample_name",
               values_to = "expression")

exp <- left_join(tophvg_exp, 
                 metadata %>% rownames_to_column("sample_name"),
                 by = "sample_name")
```

```{r}
bplot <- exp %>%
  ggplot(aes(expression, gene)) +
  geom_point(
    aes(color = condition, group = condition),
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 1),
    alpha = 0.5,
    show.legend = TRUE
  ) +
  stat_summary(
    aes(fill = condition),
    fun = median,
    geom = "col",
    position = position_dodge(1),
    alpha = 0.3
  ) +
  scale_fill_manual(values = c(
    "#6C8EBF",
    "#B07C59"
  )) +
    scale_color_manual(values = c(
    "#6C8EBF",
    "#B07C59"
  )) +
  labs(y = "Top 5 variable genes",
       x = "Gene expression") +
  ggpubr::theme_classic2()


bplot

pairs(t(norm_counts)[,tophvg], main = "Gene space")
```

## Do PCA

```{r}
# Transpose: PCA expects samples as rows
pca_res <- prcomp(t(hvg_matrix), center = TRUE, scale. = TRUE)

# Variance explained
explained_var <- pca_res$sdev^2 / sum(pca_res$sdev^2)

```

### barplot pca

```{r}
pca5 <- pca_res$x[,1:5] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("sample_name") %>% 
  pivot_longer(-sample_name,
               names_to = "PC",
               values_to = "value")

pca5 <- left_join(pca5, 
                 metadata %>% rownames_to_column("sample_name"),
                 by = "sample_name")
```


```{r}
bplot_pca <- pca5 %>%
  ggplot(aes(value, PC)) +
  geom_point(
    aes(color = condition, group = condition),
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 1),
    alpha = 0.5,
    show.legend = TRUE
  ) +
  stat_summary(
    aes(fill = condition),
    fun = median,
    geom = "col",
    position = position_dodge(1),
    alpha = 0.3
  ) +
  scale_fill_manual(values = c(
    "#6C8EBF",
    "#B07C59"
  )) +
    scale_color_manual(values = c(
    "#6C8EBF",
    "#B07C59"
  )) +
  labs(y = "Top 5 Principal Componenets",
       x = "Value") +
  ggpubr::theme_classic2()


bplot_pca
```






Show loadings of main PCA of each genes

what genes contributes to first PCs?

Biplot PCA

```{r}
require(factoextra)

var_explained <- (pca_res$sdev^2) / sum(pca_res$sdev^2)

pca_plot <- fviz_pca_biplot(pca_res,
                            habillage = metadata$condition,
                            geom = "point",
                            invisible = c("quali", "var"),
                            pointsize = 4,
                            title = ""
)
```

```{r}
pca_biplot <- fviz_pca_biplot(
  pca_res,
  habillage = metadata$condition,
  geom = "point",
  invisible = c("quali"),
  pointsize = 4,
  title = "",
  repel = T,
  select.var = list(contrib = 10)  # 👈 top 10 variables
)
```


```{r}
# Contributions of variables to PC1
contrib1 <- fviz_contrib(pca_res, choice = "var", axes = 1, top = 10)
# Contributions of variables to PC2
contrib2 <- fviz_contrib(pca_res, choice = "var", axes = 2, top = 10)

ggpubr::ggarrange(plotlist = list(contrib1, contrib2), ncol = 1)
```

```{r}
fviz_screeplot(pca_res, addlabels = TRUE)
```


```{r}
var <- get_pca_var(pca_res)
head(var$contrib)
```

# 5. single-cell annotation

## load datasets
Load and simplify Bassez dataset

```{r}
require(scGate)
require(Seurat)
require(dplyr)
require(tidyr)
require(ggplot2)
require(patchwork)
```


```{r}
sc_path <- "~/Documents/Projects/Demos/HiTME_CaseStudies/input/bassez_3patients.rds"
bassez <- readRDS(sc_path)
# one patient
bassez <- bassez[,bassez$patient_id == "BIOKEY_13"]

# ONLY 100 CELLS PER T_cell, B_cell, and Cancer_cell
set.seed(22)

# Define number of cells per category
n_cells <- 100
sel_ct <- c("T_cell", "Fibroblast", "Endothelial_cell")

# Randomly sample cells per category
cells_to_keep <- bassez@meta.data %>%
  tibble::rowid_to_column("cellid") %>% 
  filter(cellType %in% sel_ct) %>% 
  group_by(cellType) %>%            # replace 'celltype' with your column
  sample_n(size = min(n(), n_cells)) %>%   # if category < 100, take all
  pull(cellid)                    # extract cell names

# Subset Seurat object
bas <- subset(bassez, cells = cells_to_keep)

table(bas$cellType)

bas$annotation <- factor(bas$cellType,
                         levels = sel_ct,
                         labels = c("T_cell", "Fibroblast", "Endothelial"))
rm(bassez); gc()
```

## Proceessing & PCA

```{r}

bas <- NormalizeData(bas)

bas <- FindVariableFeatures(bas, 
                           selection.method = "vst", 
                           nfeatures = 1000)
bas <- ScaleData(bas, features = VariableFeatures(bas))
bas <- RunPCA(bas, features = VariableFeatures(bas))

DimPlot(bas, reduction = "pca", group.by = "annotation") 

bas <- RunUMAP(bas, dims = 1:20)

DimPlot(bas, reduction = "umap", group.by = "annotation") 
```
### pca
```{r}
pca_sd <- bas[["pca"]]@stdev         # standard deviations of PCs
pca_var <- pca_sd^2                  # variance per PC
pca_var_explained <- round(100 * pca_var / sum(pca_var), 1)  # % variance

pcaplot_sc <- DimPlot(bas, reduction = "pca") +
  scale_color_manual(values = "black") +
  xlab(paste0("PC1 (", pca_var_explained[1], "%)")) +
  ylab(paste0("PC2 (", pca_var_explained[2], "%)")) +
  NoLegend()
```
## markers

```{r}
markers_list <- list(
  T_cells       = c("CD3D", "CD3E"),
  Fibroblasts   = c("DCN", "COL1A1"),
  Endothelial   = c("PECAM1", "VWF"),
  Epithelial    = c("EPCAM", "KRT18")
)
```

## violin plot
```{r}
marker_lookup <- unlist(markers_list)
# unlist produces a vector of genes; we need names = genes, values = cell type
marker_lookup <- setNames(rep(names(markers_list), lengths(markers_list)),
                          unlist(markers_list))

norm_mat <- bas@assays$RNA@data[unlist(markers_list), ] %>% 
  as.matrix() %>% 
  t() %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("cellid") %>% 
  pivot_longer(
    -cellid,
    names_to = "gene",
    values_to = "expression"
  ) %>%
  mutate(celltype_marker = marker_lookup[gene])


vln <- norm_mat %>% 
  ggplot(aes(expression, gene, fill = celltype_marker, color = celltype_marker)) +
  geom_jitter(alpha = 0.5,
              width = 0.1) +
  geom_violin(alpha = 0.3) +
  ggpubr::theme_classic2()

vln
```

## heatmap

```{r}
require(pheatmap)
phat_mat <- bas@assays$RNA@data[unlist(markers_list), ] %>% 
  as.matrix()

rowannot <- as.data.frame(marker_lookup)
names(rowannot) <- "celltype_marker"

pheatmap(phat_mat,
         scale = "row",
         breaks = seq(-4,4,0.1),
         cluster_rows = F,
         show_colnames = F, 
         annotation_row = rowannot)
```


## feature plots

```{r}
fpl <- lapply(markers_list, function(x) x[[1]]) %>% unlist()
FeaturePlot(bas,
            features = fpl,
            reduction = "pca",
            order = T)
```

# barplots

```{r}
picked_cell <- c("BassezA_2021_33958794_BIOKEY_13_Pre_AAACGGGGTTCATGGT-1",
                 "BassezA_2021_33958794_BIOKEY_13_Pre_GCTTGAAGTCAGATAA-1",
                 "BassezA_2021_33958794_BIOKEY_13_Pre_TTTACTGTCAGTTGAC-1")


barplot_cell <- lapply(picked_cell, function(pc){
  tit <- strsplit(pc, "Pre_")[[1]][2]
  
  norm_mat %>% 
    filter(cellid == pc) %>% 
    ggplot(aes(expression, gene, fill = celltype_marker, color = celltype_marker)) +
    geom_col(alpha = 0.3) +
    labs(title = tit) +
    ggpubr::theme_classic2()
})
  
ggpubr::ggarrange(plotlist = barplot_cell,
                  nrow = 1, common.legend = T)

```




# Key messages

Key messages:

- Bioinformatics exists because biological data outgrew manual analysis.
- Method development is about making assumptions explicit and building reliable tools.
- High-dimensional data requires careful representations; PCA is a first lens.
- scRNA-seq annotation is a major applied problem with active method development.
- AI is promising, but must be validated and interpreted carefully.

```{r session-info, echo=FALSE}
# Minimal reproducibility info (helps if knitting fails on another machine)
sessionInfo()
```

# Appendix: slide-friendly prompts

Use these prompts to build slides from this Rmd:

- "Too much data" → timeline of sequencing and data growth
- Curse of dimensionality plot (dimension vs distance ratio)
- PCA toy example scatter (PC1 vs PC2)
- Marker heatmap toy example
- scGate / ProjecTILs: concept diagram (data → signatures/reference → labels)
