---
title: "Why Bioinformatics? Developing Computational Methods to Analyze Biological Data"
author: "Josep Garnica Caparros"
date: "27 March 2026"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)
set.seed(1)
```

**Rendering note**: knitting to HTML requires **Pandoc** (bundled with RStudio, or install via Quarto / system package manager).

# Audience and goals

**Audience**: mostly 3rd-year Biomedical Sciences bachelor students (plus some MSc).

**Goals of this talk**:

- Understand *why* bioinformatics is needed (the “too much data” problem).
- See how computational methods turn data into biological insight.
- Learn the intuition behind a few core ideas:
  - high-dimensional data and the curse of dimensionality
  - dimensionality reduction (PCA) as a tool for interpretation
  - cell type annotation in single-cell RNA-seq
- Understand what it means to *develop* a method/tool (assumptions → algorithm → validation → usability).

# 1. My story: how I ended up doing bioinformatics

(Short personal narrative; adapt live.)

- Started in wet lab work.
- Generated transcriptomic data, which *accumulated faster than we could analyze it*.
- During COVID lockdown: started learning bioinformatics to analyze my own data.
- Realized that good tools matter because they make analysis:
  - reproducible
  - faster
  - more interpretable
  - more accessible for others

If you want a short CV version, see: `docs/CV_JosepGarnicaCaparros_short.txt`.

## The key message

Bioinformatics is not “coding for coding’s sake”. It is **problem-driven**:

- there is a biological question
- data is generated to answer it
- computational methods are needed to extract the answer reliably

# 2. The first bioinformatics “crisis”: too much biological data

A classic example: **protein sequences** in the 1960s–70s.

- As soon as many sequences became available, humans could no longer compare them by eye.
- We needed algorithms to:
  - align sequences
  - detect similarity
  - infer evolutionary relationships
  - identify conserved motifs/domains

## A minimal framing of method development

When you build a computational method, you typically decide:

1. **Input**: what data do we accept?
2. **Representation**: how do we encode it?
3. **Objective**: what do we optimize (similarity, likelihood, classification accuracy)?
4. **Algorithm**: how do we compute it efficiently?
5. **Validation**: how do we know it works?
6. **Packaging**: how do others use it? (docs, defaults, error messages)

# 3. High-dimensional data and the curse of dimensionality

Modern biology often measures **many variables per sample**.

Examples:

- bulk RNA-seq: ~20,000 genes per sample
- single-cell RNA-seq: ~20,000 genes per cell (sparse counts)
- proteomics / metabolomics: thousands of features

This leads to two practical issues:

- **Distance becomes less informative** (everything looks similarly far away).
- **Models overfit** easily unless we have strong assumptions, regularization, or lots of data.

## Intuition: distances in high dimension

Let’s simulate points in increasing dimensions and look at how the ratio between the nearest and farthest neighbor distances changes.

# ---
# PLOT: Curse of dimensionality (dimension vs distance ratio)
# ---
```{r curse-of-dimensionality, fig.cap="Curse of dimensionality: distances concentrate", fig.path="figs/", fig.width=7, fig.height=4}
# Simulate n points in [0,1]^d and compute pairwise distances
simulate_ratio <- function(n = 400, d = 2) {
  x <- matrix(runif(n * d), nrow = n)
  # sample a subset of reference points to keep it fast
  refs <- sample(seq_len(n), size = min(30, n))
  ratios <- vapply(refs, function(i) {
    di <- sqrt(rowSums((x - matrix(x[i, ], n, d, byrow = TRUE))^2))
    di <- di[di > 0]
    min(di) / max(di)
  }, numeric(1))
  mean(ratios)
}

dims <- c(2, 5, 10, 20, 50, 100, 200)
ratios <- sapply(dims, function(d) simulate_ratio(n = 500, d = d))

df <- data.frame(d = dims, ratio = ratios)

png("figs/curse_of_dimensionality.png", width=700, height=400)
plot(df$d, df$ratio,
  type = "b", pch = 19,
  xlab = "Dimension (d)", ylab = "Average (nearest / farthest) distance",
  main = "Curse of dimensionality: distances concentrate")
dev.off()
```

**Interpretation**: as $d$ grows, the nearest neighbor is not that much closer than the farthest neighbor. That makes clustering, kNN, and other distance-based ideas harder unless we reduce dimension or use better representations.

# 4. PCA: a first tool to make high-dimensional data interpretable

**Principal Component Analysis (PCA)** finds directions (linear combinations of features) that capture the largest variance.

Why it’s useful:

- compresses many features into a few “summary axes”
- denoises (often)
- reveals structure (batches, conditions, cell states)

## PCA on a toy gene-expression-like dataset

We’ll simulate two groups with subtle correlated shifts.

# ---
# PLOT: PCA toy example (PC1 vs PC2)
# ---
```{r pca-toy, fig.cap="PCA separates subtle multigene shifts", fig.path="figs/", fig.width=7, fig.height=4}
# Simulate gene-expression-like data: samples x genes
n <- 120
p <- 200

# two groups
group <- rep(c("A", "B"), each = n/2)

# base expression
X <- matrix(rnorm(n * p), nrow = n)

# add a low-rank signal affecting a subset of genes
signal_genes <- 1:30
X[group == "B", signal_genes] <- X[group == "B", signal_genes] + 0.8

# scale genes (as typical before PCA)
X_scaled <- scale(X)

pc <- prcomp(X_scaled)

png("figs/pca_toy.png", width=700, height=400)
plot(pc$x[, 1], pc$x[, 2],
  col = ifelse(group == "A", "steelblue", "tomato"),
  pch = 19,
  xlab = "PC1", ylab = "PC2",
  main = "PCA separates subtle multigene shifts")
legend("topright", legend = c("A", "B"), col = c("steelblue", "tomato"), pch = 19)
dev.off()
```

## What PCA does and does not do

- PCA is **not** a classifier.
- PCA axes are **not automatically biological**; interpretation requires checking loadings and known biology.
- PCA is sensitive to scaling and to strong technical effects (batch).

# 5. scRNA-seq cell type annotation: from counts to cell identities

In single-cell RNA-seq, a core task is: **assign each cell a label** (e.g., T cell, B cell, monocyte; and subtypes).

## Why annotation is hard

- Data are sparse (many zeros)
- Cell states are continuous (not discrete)
- Marker genes vary across tissues, activation states, and technologies
- Batch effects and ambient RNA can confuse signatures

## Three broad strategies

1. **Marker-based** (manual): look at canonical marker genes.
2. **Reference mapping**: map your data onto a labeled reference atlas.
3. **Gating/signature-based**: automated “rules” with gene sets (analogous to flow cytometry gating).

## Two tools: scGate and ProjecTILs

- **scGate** (Carmona lab): signature-based gating for scRNA-seq.
  - Repo: https://github.com/carmonalab/scGate
  - Intuition: use curated markers and a gating model to separate cell types.

- **ProjecTILs** (Carmona lab): projects T cells onto a reference T cell atlas.
  - Repo: https://github.com/carmonalab/ProjecTILs
  - Intuition: map new data into a shared “state space” to transfer labels.

### What method development looks like here

- define robust signatures / reference states
- encode biological knowledge into a computational procedure
- validate across datasets and technologies
- make it usable (documentation, defaults, tutorials)

## (Illustrative) marker heatmap for annotation thinking

This chunk generates a *toy* marker heatmap you can use to explain the concept.

# ---
# PLOT: Toy marker heatmap (cell types x markers)
# ---
```{r marker-heatmap-toy, fig.cap="Toy marker expression by cell type", fig.path="figs/", fig.width=7, fig.height=4}
# Toy expression for 6 markers across 3 cell types
markers <- c("CD3D", "TRAC", "MS4A1", "CD79A", "LYZ", "S100A9")
celltypes <- c("T cell", "B cell", "Myeloid")

M <- matrix(c(
  2.5, 2.2, 0.2, 0.1, 0.3, 0.2,  # T cell
  0.2, 0.1, 2.7, 2.4, 0.3, 0.2,  # B cell
  0.2, 0.1, 0.2, 0.1, 2.8, 2.6   # Myeloid
), nrow = 3, byrow = TRUE)
colnames(M) <- markers
rownames(M) <- celltypes

png("figs/marker_heatmap_toy.png", width=700, height=400)
op <- par(mar = c(5, 6, 3, 1))
image(t(M[nrow(M):1, ]), axes = FALSE, col = heat.colors(50),
  main = "Toy marker expression by cell type")
axis(1, at = seq(0, 1, length.out = ncol(M)), labels = colnames(M), las = 2)
axis(2, at = seq(0, 1, length.out = nrow(M)), labels = rev(rownames(M)), las = 2)
par(op)
dev.off()
```

# 6. Current challenges and the future: AI in bioinformatics

AI is powerful, but it is not magic. Useful framing:

- **AI is a tool for pattern recognition**, often very strong at prediction.
- In biology, we often need:
  - interpretability
  - uncertainty quantification
  - causal reasoning
  - generalization across cohorts and technologies

## Advantages

- Can model nonlinear relationships.
- Can integrate multiple modalities (RNA + protein + images).
- Can scale to very large datasets.

## Limitations / risks

- Data leakage and confounding.
- Lack of interpretability.
- Distribution shift (model trained on one dataset fails on another).
- “Hallucinated biology”: plausible-looking but incorrect explanations.

## A practical takeaway

When you evaluate a new method (AI or not), ask:

- What assumptions does it make?
- What data was it trained/validated on?
- What are failure modes?
- Does it provide uncertainty / confidence?

# 7. Key messages

Key messages:

- Bioinformatics exists because biological data outgrew manual analysis.
- Method development is about making assumptions explicit and building reliable tools.
- High-dimensional data requires careful representations; PCA is a first lens.
- scRNA-seq annotation is a major applied problem with active method development.
- AI is promising, but must be validated and interpreted carefully.

```{r session-info, echo=FALSE}
# Minimal reproducibility info (helps if knitting fails on another machine)
sessionInfo()
```

# Appendix: slide-friendly prompts

Use these prompts to build slides from this Rmd:

- "Too much data" → timeline of sequencing and data growth
- Curse of dimensionality plot (dimension vs distance ratio)
- PCA toy example scatter (PC1 vs PC2)
- Marker heatmap toy example
- scGate / ProjecTILs: concept diagram (data → signatures/reference → labels)
